---
title: "Generational Dynamics, Gender and Politics in the US"
subtitle: "Predicting 2020 US Presidential Voting Preferences"
author: 
  - Hritik Shukla
thanks: "Code and data are available at: https://github.com/hritikshuklas/vote-preferences-generations"
date: today
date-format: long
abstract: "This study investigates voting patterns in the 2020 US presidential election, analyzing how political preferences are influenced by a person's gender and the generation they're born in. With the help of exploratory data analysis, logistic regression and data from the 2020 CES Common Consent dataset, it is observed that younger generations tend to favor Biden over Trump. As generations get older, Biden loses more and more support. It is also found that women tend to favor Biden more heavily than the men of the same generation as them."
format: pdf
output:
  pdf_document: default
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(rstanarm)
library(arrow)
library(ggplot2)
library(knitr)
library(modelsummary)
```

```{r}
#| include: false
#| warning: false
#| message: false

raw_data <- read_csv("../data/raw_data/ces2020.csv")

analysis_data <- read_parquet("../data/analysis_data/cleaned_data.parquet")

analysis_data <- analysis_data |>
  mutate(
    voted_for = factor(
      voted_for,
      levels = c("Biden", "Trump")
    ),
    gender = factor(
      gender,
      levels = c("Female", "Male")
    ),
    generation = factor(
      generation,
      levels = c(
        "Silent Generation",
        "Baby Boomer",
        "Generation X",
        "Millennial",
        "Generation Z"
      )
    )
  ) |>
  select(gender, birthyear, generation, voted_for)

model <- readRDS(file="../models/generation_gender_model.rds")
```

# Introduction {#sec-intro}

*‘If you’re not a liberal when you’re 25, you have no heart. If you’re
not a conservative by the time you’re 35, you have no brain.’*

Although commonly, but incorrectly attributed to Winston Churchill, this
quote summarizes a common belief held in our society - a person
generally starts out in life with a left leaning political compass, and
as they get older, their beliefs transform into more right leaning
ideologies. It is also a popular belief that newer, younger generations
tend to be more inclusive, open and liberal in their ideologies
(@washingtonpost), and this only intensifies as further generations get
old enough to participate in the political landscape of the US.

This paper aims to find out whether these generalizations hold true by
observing the 2020 US Presidential Elections. It is important to note
that 2020 was a tumultuous year for the entire world, it first year of
COVID-19 when not much was known about the disease. President Trump, who
was in power before the elections, was greatly criticized for his poor
handling of pandemic and lockdown, which lead to massive change in
sentimenets amongst the citizens of the country (@trumpdisapproval). As
a result, with highly charged sentiments, this election saw the highest
voter turnout of any presidential election held in the US in the 21st
century (@2020voterturnout). Another interesting factor to consider is
that this was the first election where Generation Z could participate.
Furthermore, they held highest voter turnout of any generation
(@genzturnout). Therefore, it would be even more interesting to observe
if these long accepted norms held up in these extraordinary conditions,
under situations which would not be present under any other normal
circumstances - the first election of its kind in recent memory.

The remainder of this paper is structured as follows:

-   @sec-data explores the dataset and the variables within it used for
    the study

-   @sec-model explains our models setup and our assumptions going into
    the study

-   @sec-results explores our findings in detail

-   @sec-discussion evaluates these findings in the context of our
    reality

I used R (@citeR), along with multiple packages to aid in the data
analysis and modelling. More particularly, the packages tidyverse
(@tidyverse), dplyr (@citeDplyr), tidyr (@citeTidyr), dataverse
(@dataverse) and arrow (@citeArrow) were used for data acquisition,
testing and cleaning. The package rstanarm (@citeRstanarm) was used for
modelling, and ggplot2 (@citeGgplot2), knitr (@citeKnitr) and
modelsummary (@citeModelsummary) were used for data visualization.

# Data {#sec-data}

The final release of the 2020 CES Common Consent dataset by (@dataset)
was used for this study, acquired through the dataverse package
(@dataverse). This dataset was created from a survey conducted by
YouGov, an internet-based data analytics firm based in the UK. Random
sample methodology was used to survey 61,000 adults over the internet
from September to October 2020 (@datasheet). For our model, the "gender"
(@sec-data-gender) and "CC20_410" (@sec-data-voted-for) variables were
used from this dataset directly, and the "birthyr" variable was used to
construct the "generation" variable (@sec-data-generation).

## Data Cleaning {#sec-data-cleaning}

```{r}
#| label: tbl-raw-data-preview
#| tbl-cap: Raw Dataset Preview
#| echo: false

kable(head(raw_data, 10), booktabs=TRUE, format="latex")
```

The "votereg", "CC20_410", "gender" and "birthyr" variables from CCES
2020 dataset (as seen in @tbl-raw-data-preview) were selected initially.

As the first step, any voters who weren't registered to vote were
removed from our data. The "votereg" variable was used for this process,
and then was subsequently removed from our finalized data as it served
no purpose for our model. Next, "CC20_410", which records who the
respondent voted for for President of the Untied States, was used to
filter out respondents who voted for Joe Biden or Donald J. Trump, as we
are interested in peoples' preference towards the Democrats or the
Republicans. Any respondents who were born outside the years 1928 to
2012 were removed from the survey, as there wouldn't be enough data for
those generations to make any statistical inferences from them.

Next, any responses which were missing any of the these variables we
needed were removed during cleaning. This omission culled down our total
responses from 61,000 to 43,540. However, we still have more than enough
data to derive meaningful results.

Some variables were renamed to be more human-readable - "CC20_410" was
renamed to "voted_for" and "birthyr" was renamed to "birthyear". The
values for categorical variables (such as "gender" and "voted_for")
within the dataset were cleaned to represent their more meaningful,
intended values. The questions corresponding to these variables in the
survey have a select number of options for the subject to choose from.
These responses are recorded in the dataset as a number, where each
number corresponds to a relevant response - for example, for the gender
variable, a response of 1 means female and 2 means male. During
cleaning, these numbered responses were replaced by their true values.

The end result of the cleaning process leaves the dataset shown in
@tbl-cleaned-data-preview.

```{r}
#| label: tbl-cleaned-data-preview
#| tbl-cap: Cleaned Dataset preview
#| echo: false

col_names <- c("Gender", "Birthyear", "Generation", "Voted For")
kable(head(analysis_data, 10), format="latex", booktabs=TRUE, col.names=col_names)
```

## Gender {#sec-data-gender}

```{r}
#| label: fig-gender-histogram
#| fig-cap: Gender Distribution
#| echo: false

# Create a histogram
ggplot(analysis_data, aes(x = gender)) +
  geom_bar(width=0.3) +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5) +
  labs(x = "", y = "Number of respondents") +
  ylim(0, 27000) +
  theme_minimal()
```

The gender variable stores the responses of the respondent to a question
which asks them to choose between "Male" and "Female" as their only two
options to describe their gender. This question would have been better
phrased if it were asking for biological sex assigned at birth rather
than gender based on the options provided to the respondents. Even
though the survey asks for the respondent's sexual orientation later,
the phrasing of this question alone could be a reason for people to not
submit their survey responses as they don't identify with the given
options which may lead to under-representation of people belonging to
this demographic in the dataset.

In @fig-gender-histogram, we can see the number of respondents who
identified as male and female. According to @genderratio, the United
States had a gender ratio of 97.14 males to 100 females in 2020. Our
data consists of 24,295 females and 19,245 male - the gender ratio
represented here is considerably below the national average at the time
of the survey - this might introduce some bias into our data.

## Voted For (CC20_410) {#sec-data-voted-for}

```{r}
#| label: fig-vote-histogram
#| fig-cap: Vote Distribution
#| echo: false

# Create a histogram
ggplot(analysis_data, aes(x = voted_for)) +
  geom_bar(width=0.3) +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5) +
  labs(x = "", y = "Number of respondents") +
  ylim(0, 28000) +
  theme_minimal()
```

The CC20_410 variable, renamed to "voted_for", records who the
respondent voted for in the 2020 Presidential Elections. The respondents
were given options other than Joe Biden and Donald Trump for this
question, such as "Other", "I dd not vote", "Not Sure", etc, but these
options were removed as they weren't needed for our purposes.

In @fig-vote-histogram, we can see that Biden has an overwhelming
majority over Trump in terms of popularity, with around a 6,000 vote
difference between the two.

## Generation (and birthyear) {#sec-data-generation}

```{r}
#| label: fig-generation-histogram
#| fig-cap: Generation Distribution
#| echo: false

# Create a histogram
ggplot(analysis_data, aes(x = generation)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5) +
  labs(x = "", y = "Number of respondents") +
  ylim(0, 20000) +
  theme_minimal()
```

The generation variable was created from the "birthyr" variable in the
dataset, where respondents recorded their year or birth. Each respondent
was assigned a "generation" according to their year of birth. The
generation breakpoints used for this paper are defined by Pew Research
Center (@pewgenerations) as follows:

-   Silent Generation: 1928 - 1945

-   Baby Boomer: 1946 - 1964

-   Generation X: 1965 - 1980

-   Millennial: 1981 - 1996

-   Generation Z: 1997 - 2012

Any respondents whose year of birth falls outside of these year ranges
were removed from the dataset. These numbers fall in line with the voter
turnout

In @fig-generation-histogram, we can see that Baby Boomers show the
largest representation of all groups. Generation X and Millennial have
similar number of respondents, and Silent Generation and Generation Z
have the lowest number of respondents. This distribution falls in line
with the voter turnout by age statistic reported by Pew Research Center
(@pewvoterturnout). Note that even though Generation Z had the highest
voter turnout for all generations (@pewgenerations) and their numbers
look low compared to others, it falls in line with expected numbers as
the oldest Generation Z respondent who would've been able to vote
would've been born in 2002, leaving a considerable population of this
generation unable to vote.

# Model {#sec-model}

We will be using logistic regression to model our data, where our
outcome variable would be whether a respondent prefers Biden as the
presidential candidate. Gender and generation will be used as predictors
our outcome variable.

## Model set-up

Define:

-   $y_i$ is the political preference of the respondent and equal to 1
    if Biden, and 0 if Trump

-   $\mbox{gender}_i$ is the gender of the respondent

-   $\mbox{generation}_i$ is the generation of the respondent

```{=tex}
\begin{align}  y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\ \mbox{logit}(\pi_i) &= \beta_0 + \beta_1 \times \mbox{gender}_i +  \beta_2 \times \mbox{generation}_i\\ \beta_0 &\sim \mbox{Normal}(0, 2.5) \\ \beta_1 &\sim \mbox{Normal}(0, 2.5) \\ \beta_2 &\sim \mbox{Normal}(0, 2.5) \end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of
@rstanarm. We use the default priors from `rstanarm`.

### Model justification

Logistic regression is employed for this model since our variable of
interest can be constructed as a binary outcome variable - respondent
prefers Biden, respondent doesn't prefer Biden (prefers Trump). Logistic
regression is well suited for situations where the outcome variable
represents two mutually exclusive categories and its probability is
based on a set of predictor variables (here, gender and generation).

According to the common beliefs in our society as established in
@sec-intro, we expect to see a positive relationship in the younger
generations, who we expect to be more left-leaning and therefore be more
favorable towards Biden. And vice versa, we expect to see a negative
relationship in the older generations, who we expect to be more right
leaning, and therefore, more conservative. However, due to the large
dissatisfaction in the majority of US citizens in 2020 due to how
President Trump handled the COVID pandemic (@trumpdisapproval), this
skew might not be as great as it could be for other elections.
Regardless, we expect to see a linear relationship between a person's
generation and preference for Biden - younger generational cohorts
leaning more towards voting for Biden.

The relationship between gender and political preference is a bit more
complicated. According to articles from the Conversation
(@conversationwomen) and the Gallup (@gallupwomen), women tend to vote
more conservative than men before 2017, after which this trend flipped
on its head and women are began to vote more liberally.

# Results {#sec-results}

Our results are summarized in.

```{r}
#| label: fig-pres-pref-gender-generation
#| fig-cap: Distribution of presidential preferences, by gender, and by generation
#| echo: false

# Create a histogram
analysis_data |>
  ggplot(aes(x = generation, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(gender)) +
  theme_minimal() +
  labs(
    x = "Generation",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  theme(legend.position = "bottom") + 
  theme_minimal()
```

@fig-pres-pref-gender-generation Compares the political preferences
between respondents of different generations, separated by gender. The
voting behaviour on this histogram closely resembles most of our
assumptions, but there are some interesting observations which weren't
predicted.

First big observation is the fact that as generational cohorts get
younger, the difference in votes cast to Biden and Trump increase in the
favor of Biden across both genders, showing that younger voters tend to
be more liberally aligned politically. This phenomenon is exaggerated in
the female voter base of each generation - the difference between the
votes cast to Biden and Trump are much greater compared to those in
males.

Between the two genders, female voters showed a much higher discrepancy
in their voting preferences in the favour of Biden. This falls in line
with the recent trends of women voting more liberal in the recent years.

Note that these observations are further supported by the coefficients
of our fitted logistic model, as seen in @tbl-model-summary:

```{r}
#| label: tbl-model-summary
#| tbl-cap: Model Summary
#| echo: false
#| warning: false

modelsummary(
  list(
    "Support Biden" = model
  ),
  statistic = "mad"
)

```

Note that for our model, we set Biden as our reference level for
voted_for, Silent Generation as our reference level for generation, and
Female as our reference level for gender. This means that the
coefficients of other generations are relative to the Silent Generation,
which we have seen from the histogram tend to favor Trump over Biden.
Moreover, R treats "Biden" as our failure case and "Trump" as our
success case, which implies that negative coefficients imply a positive
increase in odds for Biden. Conversely, a positive coefficient imply a
decrease in odds for voting Biden, i.e., an increase in odds of voting
for Trump. Keeping these key pieces of information in mind, we proceed
to our model's results.

Our model arrived at the intercept of 0.051 with a standard error of
0.035. This means that we can say with fairly high accuracy, that when
all predictors are set to 0, i.e, for the female demographic of the
Silent Generation, Trump's log odds for getting a vote is 0.051, which
is about 51% probability - this falls in line with our observations in
@fig-pres-pref-gender-generation.

Keeping our reference level of Silent Generation and Biden in mind, we
now observe the coefficients for different generations:

-   Baby Boomers have a coefficient of -0.422 with a standard error of
    0.036

-   Generation X has a coefficient of -0.597 with a standard error of
    0.040

-   Millennials have a coefficient of -1.116 with a standard error of
    0.041

-   Generation Z have a coefficient of -1.393 with a standard error of
    0.070

Notice that the coefficients tend to decrease further and further as the
generational cohorts become younger, implying that Biden tends to gain
more favor as the generational cohorts get younger - keeping in line
with our hypothesis. Moreover, the standard error for each of these
generations is very small, implying a high precision to these
predictions made by our model.

Finally, we observe the coefficient for male gender to be 0.328 with a
standard error of 0.020, which when compared to the female gender favors
Trump over Biden. This implies that male respondents showed an increase
of 0.328 log odds in voting for Trump when compared to female
respondents. And due to our very low standard error, we can say with
high confidence that male respondents have a higher chance of voting for
Trump compared to female respondents, which also agrees with our
hypothesis.

# Discussion {#sec-discussion}

## What was done {#sec-first-point}

It is an age old belief that as a person starts out with a left leaning
political compass and as they get older, their ideology slowly shifts
over to the right. The purpose of this paper was to test this belief by
gaining a better understanding of the voting preferences (between Joe
Biden and Donald J. Trump) different generations had in the 2020 US
Presidential Elections, based on the generation they belonged to, as
well as the gender of the members of these generations. This was done by
obtaining data from the 2020 CES Common Consent dataset, which contained
survey responses from over 61,000 adults surveyed over the internet.
Initial exploratory data analysis suggested that younger generations
tended to favor Biden over Trump, and women tended to favor Biden more
heavily than men of the same generation. Logistic regression modelling
was then used to formalize this discovery, and it showed a clear trend
which agreed with our initial hypothesis - as generational cohorts get
younger, the odds of favoring Biden increase.

## "Younger Generations vote Liberally; Older Generations vote Conservative"

Our main hypothesis of the study was proven to be true from our
findings - younger generations vote liberally; older generations vote
conservative. From both exploratory analysis and our model, it was found
that younger generations, such as Millennial and Generation Z, tend to
favor supporting liberal candidates (Biden in this case). As
generational cohorts got older, their liberal support slowly decreased
until eventually the generation as a whole tended to favor the
conservative candidate (Trump). Do note that even though Trump's
favorability increased as generations got older, Biden still held
majority votes across all generations other than the Silent Generation,
who are now a minority in the American voter base. This might be
explained by Trump's growing unpopularity in 2020 from the way he
handled the COVID-19 pandemic, and over his term as the President as a
whole. What's more interesting is that even though only a part of
Generation Z participated in the election, and it was the first election
for the generation as a whole, they fit in the trend perfectly, showing
overwhelming support for Biden in their votes compared to other
generations, even more so than Millennials.

## "Women Vote more Liberally than Men"

A number of studies published in the last few years (@gallupwomen,
@conversationwomen) suggested that women were voting more liberally than
before, and this was observed in our own findings as well. Against the
traditional beliefs that women tended to vote more conservative than
men, it was found that women were tended to favor the liberal candidate
much greater than the conservative candidates across all generations.

## Weaknesses and next steps

Some weaknesses of this model arrive from the data that was used for
this study - the 2020 CES Common Consent dataset. Even though it had
over 61,000 respondents, over 6,000 respondents either didn't register
to vote or had no idea if they were registered at all. Moreover, after
data cleaning was completed, we were left with 43,540 entries in the
dataset - over a third of the data was lost. Furthermore, in these
43,540 entries, the ratio of male to female entries weighed greatly in
favor of females, and was much higher than the national gender ratio in
2020.

Due to the nature of the survey being online, a significant portion of
the US's demographics was left out of the pool. It is estimated that in
2020 about 13% of the US population did not have access to the internet
(@statista), which leads to the exclusion of about 43 million citizens
being excluded from the survey with no representation at all.

It would be interesting to study how different generations tend to vote
in the previous US presidential elections (which occurred under more
normal circumstances) as well, especially the cases where a new
generation was added to the voter pool of the nation, similar to 2020.
Another interesting avenue would be to study voting trends amongst women
over the last few decades, as the scope of this paper limits us from
exploring exactly how great of a shift occurred in 2017.

\newpage

\appendix

# Appendix {.unnumbered}

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check. In @fig-ppcheckandposteriorvsprior-2 we compare the posterior
with the prior.

```{r}
#| eval: true 
#| echo: false 
#| message: false 
#| warning: false 
#| label: fig-ppcheckandposteriorvsprior 
#| layout-ncol: 2 
#| fig-cap: Examining how the model fits, and is affected by, the data 
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(model) + 
  theme_classic() + 
  theme(legend.position ="bottom")

posterior_vs_prior(model) + 
  theme_minimal() + 
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position ="bottom") + 
  coord_flip()
```

## Diagnostics

```{r}
#| label: fig-model-cred-interval 
#| fig-cap: Credible Intervals
#| echo: false 
#| warning: false

modelplot(model, conf_level = 0.9) + labs(x = "90 per cent credibility
interval")
```

```{r}
#| echo: false 
#| eval: true 
#| message: false 
#| warning: false
#| label: fig-model-trace-rhat 
#| fig-subcap: ["Trace plot", "Rhat plot"] 
#| layout-ncol: 2

plot(model, "trace")

plot(model, "rhat")

```

\newpage

# References
